A Bidirected String Graph Model for Genome Assembly

### 2012-12-11 presentation/paper.tex
paper: mapping contained reads

The contained reads cannot be completely discarded because they are needed for a
later step (see \Section{sec:mapping_contained_reads}).

@@

We would like to use hints from the original data to restrict the number of
times each bidirected edge may be traversed.  This can be done using a
calculation called the \KeyTerm{A-statistic}, or \KeyTerm{arrival-rate
statistic}, which is the logarithm of the probability that an edge represents
unique sequence rather than duplicate sequence\cite{Myers2005}.  The A-statistic
$A(e)$ of an edge $e$ in a bidirected string graph is:

@@

where $n$ is the total number of reads, $G$ is the estimated genome length,
$\Delta$ is the length of the edge in bp, and $k$ is the number of reads that
support that edge.  To more precise, $k$ is the number of inner vertices in the
unbranched path corresponding to $e$ before the unbranched paths were collapsed.

Edges with a sufficiently high A-statistic are very likely to be single-copy
edges and therefore should be traversed exactly one time.  The remaining edges
may be traversed any number of times (with the exact number being determined
during the network flow calculation), except for edges that originally were
associated with one or more inner vertices, in which case we require the edge be
traversed at least one time so that the reads corresponding to the inner
vertices are used in the reconstruction of the genome (as every read was
presumably sampled from the genome, in the absence of foreign DNA).

Calculation of the A-statistic on each edge in the bidirected string graph is
implemented by the program {\tt calculate-A-statistics}.

@@

\label{sec:mapping_contained_reads}

>>>>

The contained reads cannot be completely discarded because they are needed for a
later step (see \Subsection{subsec:mapping_contained_reads}).

@@

We would like to use hints from the original data to restrict the number of
times each bidirected edge may be traversed.  This can be done using a
calculation called the \KeyTerm{A-statistic}, or \KeyTerm{arrival-rate
statistic}, which is the logarithm of the probability that an edge represents
unique sequence rather than duplicate sequence \cite{Myers2005}.  The A-statistic
$A(e)$ of an edge $e$ in a bidirected string graph is:

@@

where $n$ is the total number of reads, $G$ is the estimated genome length,
$\Delta$ is the length of the edge in bp, and $k$ is the number of reads that
support that edge.  To more precise, $k$ is the number of inner vertices in the
unbranched path corresponding to $e$ before the unbranched paths were collapsed
\cite{Myers2005}.

Edges with sufficiently high A-statistic are very likely to be single-copy edges
and therefore should be traversed exactly one time.  The remaining edges may be
traversed any number of times (with the exact number being determined during the
network flow calculation), except for edges that originally were associated with
one or more inner vertices, in which case we require the edge be traversed at
least one time so that the reads corresponding to the inner vertices are used in
the reconstruction of the genome.  This is needed because in the absence of
foreign DNA, every read was presumably sampled from the genome.

However, $G$, the estimated genome length, is actually unknown, so it must be
estimated first.  It can be estimated by producing a ``bootstrap'' genome length
by adding up the lengths of the sequences on every edge in both directions and
dividing by 2, then calculating the A-statistics using this bootstrap genome
length and identifying edges that are probably single-copy.  The A-statistics
can then be re-estimated over just these single-copy edges in order to refine
the genome length further before calculating the final A-statistic on all the
edges.  This procedure was outlined in \cite{Myers2005}.

The calculation of the A-statistic on each edge in the bidirected string graph,
including the genome length bootstrapping algorithm, is implemented by the
program {\tt calculate-A-statistics}.

@@

Because contained reads and their overlaps were removed before constructing the
assembly graph (see \Subsection{subsec:removing_contained_reads}), the
calculation of the A-statistic as described above is flawed because contained
reads need to be counted.  This is especially a problem for high copy-number
regions because those regions are likely to generate many contained reads.
Therefore, we need to map the contained reads into the graph before calculating
the A-statistics.

A contained read may easily be mapped to the vertex corresponding to a read that
contains it, but this is not sufficient because the A-statistic is calculated
over the edges, not the vertices.  Instead, we need to map each end of each the
contained read to the edge, or in general the set of edges, that contains it.

The algorithm for doing so is outlined in \cite{Myers2005} and was implemented
by the program \ProgramName{map-contained-reads}, which runs before
\ProgramName{collapse-unbranched-paths}.  Briefly, for each contained read $r$
the algorithm identifies the uncontained read $R$ that contains $r$ and has the
shortest distance from the end of $r$ to the end of $R$ (the shortest
\KeyTerm{overhang}).  This is done for each of the two directions.  The edges in
the graph starting at the vertex corresponding to that end of $R$ are then
walked backwards for the length of the overhang to find the set of edges $S$ to
which the corresponding edge of $r$ may map, and each edge in $S$ is marked as
containing $\frac{1}{|S|}$ extra inner vertices.  This mapping, therefore, makes
it possible to account for contained reads when calculating the A-statistic.

\label{subsec:mapping_contained_reads}
